One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-15 paper/appendix.tex
minor updates

\section{Relation between horizontal Lagragian displacements for
  local, low frequency $m=1$ disturbances}\label{horizontal_displacements}
Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$
and $\xi_\phi$ in the local approximation. Using the local solution to
the Poisson equation 
\begin{align}
  \delta \Phi_m = -\frac{2\pi G}{|k|} \delta\Sigma_m 
\end{align}
\citep{shu91}, the linearised azimuthal equation of motion becomes 
\begin{align} 
  - \ii\sbar \delta v_{\phi m}  + \frac{\kappa^2}{2\Omega}\delta v_{Rm} = -\frac{\ii
    m}{R\Sigma}\left(c_s^2 - \frac{2\pi G
      \Sigma}{|k|}\right)\delta\Sigma_m. 
\end{align}
Next, we replace the surface density perturbation
$\delta \Sigma_m = -\ii k \Sigma \xi_R$, and use the expressions
\begin{align}
  &\delta v_{Rm} = -\ii\sbar\xi_R,\\
  &\delta v_{\phi m} = -\ii\sbar\xi_\phi - \frac{\ii R
    \p_R\Omega}{\sbar} \delta v_{Rm}
\end{align}
\citep{papaloizou85} to obtain 
\begin{align}
  -\sbar^2\xi_\phi - 2\ii\sbar\Omega \xi_R =
  \frac{m}{kR}\left(\kappa^2 - \sbar^2\right)\xi_R, 
\end{align}
where the dispersion relation Eq. \ref{dispersion} was used. 
In the local approximation, $|kR|\gg m$ by assumption.%  Furthermore,  
% Also, for low frequency $m=1$ modes $\sbar \simeq -\Omega$, so the
% quantity $|\kappa^2 - \sbar^2| \ll \Omega^2$ in a nearly Keplerian disc
% ($\kappa\sim \Omega$). 
 Hence the RHS of this equation can be neglected. Then 
\begin{align}
  \xi_\phi \simeq -\frac{2\ii\Omega}{\sbar}\xi_R.
\end{align}
For low-frequency $m=1$ modes we have $\sbar \simeq -\Omega$, so
$\xi_\phi\simeq 2\ii\xi_R$, as used in the main text.

>>>>

\section{Relation between horizontal Lagragian displacements for
  local, low frequency \lowercase{$m=1$} disturbances}\label{horizontal_displacements}
Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$
and $\xi_\phi$ in the local approximation. Using the local solution to
the Poisson equation 
\begin{align}
  \delta \Phi_m = -\frac{2\pi G}{|k|} \delta\Sigma_m 
\end{align}
\citep{shu91}, the linearised azimuthal equation of motion becomes 
\begin{align} 
  - \ii\sbar \delta v_{\phi m}  + \frac{\kappa^2}{2\Omega}\delta v_{Rm} = -\frac{\ii
    m}{R\Sigma}\left(c_s^2 - \frac{2\pi G
      \Sigma}{|k|}\right)\delta\Sigma_m. 
\end{align}
Next, we replace the surface density perturbation
$\delta \Sigma_m = -\ii k \Sigma \xi_R$, and use the expressions
\begin{align}
  &\delta v_{Rm} = -\ii\sbar\xi_R,\\
  &\delta v_{\phi m} = -\ii\sbar\xi_\phi - \frac{\ii R
    \p_R\Omega}{\sbar} \delta v_{Rm}
\end{align}
\citep{papaloizou85} to obtain 
\begin{align}
  -\sbar^2\xi_\phi - 2\ii\sbar\Omega \xi_R =
  \frac{m}{kR}\left(\kappa^2 - \sbar^2\right)\xi_R, 
\end{align}
where the dispersion relation Eq. \ref{dispersion} was used. 
In the local approximation, $|kR|\gg m$ by assumption.%  Furthermore,  
% Also, for low frequency $m=1$ modes $\sbar \simeq -\Omega$, so the
% quantity $|\kappa^2 - \sbar^2| \ll \Omega^2$ in a nearly Keplerian disc
% ($\kappa\sim \Omega$). 
 Hence the RHS of this equation can be neglected. Then 
\begin{align}
  \xi_\phi \simeq -\frac{2\ii\Omega}{\sbar}\xi_R.
\end{align}
For low-frequency $m=1$ modes we have $\sbar \simeq -\Omega$, so
$\xi_\phi\simeq 2\ii\xi_R$, as used in the main text.

### 2014-11-15 paper/discussion.tex
minor updates

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes generally have negative
angular momentum. If the mode loses (positive) angular momentum
to the background, then we can expect instability. We demonstrate
this below.

@@

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. Here, we estimate the magnitude of this effect using
basic results from disc-planet theory \citep[see, e.g.][and references 
therein]{papaloizou07}.

Let us treat the one-arm spiral confined in $R\in[R_1,R_2]$ as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We take 
\begin{align} 
  \Phi_\mathrm{ext}
  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
\end{align}
where $M_\mathrm{ring}$ is the disc mass contained within
$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
and $\beta = R/\overline{R}$. This form of
$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
potential of an external satellite on a circular orbit
\citep{goldreich79}.

@@

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
  &\times \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-arm spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}   
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right)
\left[b_{1/2}^1(\beta_c)\right]^2   
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

These estimates are much smaller than that due to
the imposed temperature gradient as measured in the simulations
($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
Lindblad and co-rotation resonances have negligible effect on the
growth of the spiral arm in the inner disc. We confirmed this with
additional FARGO simulations with a reduced radial domain size,
setting $R_\mathrm{max}=3R_0$, which still developed the one-arm
spiral.

>>>>

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes generally have negative
angular momentum. If the mode loses (positive) angular momentum
to the background, then we can expect instability. We demonstrate
this below.

@@

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-armed spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. Here, we estimate the magnitude of this effect using
basic results from disc-planet theory \citep[see, e.g.][and references 
therein]{papaloizou07}.

Let us treat the one-armed spiral confined in $R\in[R_1,R_2]$ as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We take 
\begin{align} 
  \Phi_\mathrm{ext}
  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
\end{align}
where $M_\mathrm{ring}$ is the disc mass contained within
$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
and $\beta = R/\overline{R}$. This form of
$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
potential of an external satellite on a circular orbit
\citep{goldreich79}.

@@

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
  &\times \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}   
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right)
\left[b_{1/2}^1(\beta_c)\right]^2   
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

These estimates are much smaller than that due to
the imposed temperature gradient as measured in the simulations
($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
Lindblad and co-rotation resonances have negligible effect on the
growth of the spiral arm in the inner disc. We confirmed this with
additional FARGO simulations with a reduced radial domain size,
setting $R_\mathrm{max}=3R_0$, which still developed the one-armed
spiral.

### 2014-11-15 paper/model.tex
simplified discussion
minor updates

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Thus, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant
\citep{adams89,shu90}. Indeed, simulations including the indirect
potential, carried out in the early stages of this project, produced
similar results.

@@

In 3D, we assume vertical hydrostatic balance
\begin{align}
  0 = \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_*}{\p z} + \frac{\p
    \Phi_d}{\p z},  
\end{align}
which gives the mass density as 
\begin{align}
  \rho = \frac{\Sigma}{\sqrt{2\pi}H}Z(R,z),
\end{align}
where $Z(R,z)$ describes vertical stratification. In practice, we
numerically solve for $Z(R,z)$ by neglecting the radial self-gravity
force compared to vertical self-gravity, which reduces the equations
for vertical hydrostatic equilibrium to ordinary differential
equations. This procedure is described in \cite{lin12b}.

@@

% poisson kernal integration, softening 
% where used?
% BC
\subsection{ZEUS-MP}
ZEUS-MP  is a general-purpose finite difference
code \citep{hayes06}. We use the code in 3D spherical geometry, covering
$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
The grid is logarithmically spaced in radius and uniformly spaced in the angular
coordinates. We assume reflectional symmetry across the midplane, and
apply relfective boundary conditions at radial boundaries and the
upper disc boundary.

@@

\subsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the angular momentum flux
consisting of a Reynolds stress and a gravitational stress. Explicit
expressions for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85},
and $\bm{F}$ can be read off Eq. \ref{angflux1}.  
% \cite{papaloizou85} and \cite{lin11} respectively, and

@@

The background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momeumtum flux and the perturbations are regular.

>>>>

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant
\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
potential, carried out in the early stages of this project, produced
similar results.

@@

%In 3D, we assume vertical hydrostatic balance
The 3D disc structure is obtained by assuming vertical hydrostatic
balance  
\begin{align}
  0 = \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_*}{\p z} + \frac{\p
    \Phi_d}{\p z},  
\end{align}
which gives the mass density as 
\begin{align}
  \rho = \frac{\Sigma}{\sqrt{2\pi}H}Z(R,z),
\end{align}
where $Z(R,z)$ describes vertical stratification. In practice, we
numerically solve for $Z(R,z)$ by neglecting the radial self-gravity
force compared to vertical self-gravity, which reduces the equations
for vertical hydrostatic equilibrium to ordinary differential
equations. This procedure is described in \cite{lin12b}.

@@

% poisson kernal integration, softening 
% where used?
% BC
\subsection{ZEUS-MP}
ZEUS-MP  is a general-purpose finite difference
code \citep{hayes06}. We use the code in 3D spherical geometry, covering
$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
The grid is logarithmically spaced in radius and uniformly spaced in the angular
coordinates. We assume reflectional symmetry across the midplane, and
apply reflective boundary conditions at radial boundaries and the
upper disc boundary.

@@

\subsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the vertically-integrated angular
momentum flux consisting of a Reynolds stress and a gravitational
stress \citep{lin11b}. Explicit expressions for $\bm{\xi}$ can be found in,
e.g. \cite{papaloizou85}. 
%and $\bm{F}$ can be read off Eq. \ref{angflux1}.

% \cite{papaloizou85} and \cite{lin11} respectively, and

@@

The background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the total angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momeumtum flux.% and the perturbations are regular.
